---
title: "Getting to Know Your Data"
---

## Variables

Before we jump into the world of statistical testing, let’s take a moment to refresh our understanding of what a variable is, the different types of variables, and how we can describe them using numerical and graphical methods.

**So what is a variable?**

The Australian Bureau of Statistics defines a variable to be "any characteristic, number, or quantity that can be measured or counted."

If you’ve ever tried dieting, you might have recorded your weight on your phone over time. Or maybe you’ve tracked how long your morning commute takes so you know when to leave to arrive on time. In both cases, you are measuring something that changes, and these are examples of variables.

## Types of Variables

There are two main types of variables:

-   **Numerical**

-   **Categorical**

A **numerical variable** represents values to describe a measurable quantity. A numerical variable can be either **discrete** or **continuous**.

**Discrete numerical variables** take on distinct whole values, and are countable. For example, if we were to look at the at the number of cars in a parking lot, the number of birds in a tree, or the number of times "3" is rolled in a series of die rolls, these are exact values that can be counted and are thus discrete.

**Continuous numerical variables** can take on any value within a range of real numbers. In other words, they can be measured to any level of precision. For example, age exists on a continuous scale because, in theory, it can be measured infinitely precisely. Someone could be 10.25, 10.257, or even 10.2576326362 years old. In everyday life, we usually round age to the nearest year, but that does not change the fact that the underlying variable is continuous. Other common examples of continuous numerical variables include weight, height, and time. Can you explain why each of these variables is numerical and continuous?

A **categorical variable**, on the other hand, represents categories or labels that describe a quality or characteristic. Categorical variables can be either **nominal** or **ordinal**.

A **nominal categorical variable** includes observations with no particular order. For example, if you asked ten coworkers for their favorite movie and recorded their answers under a "movies" variable, it would be nominal because there is no inherent sequence or ranking among the movie titles.

In an **ordinal categorical variable,** there is an inherent logical order to the observations. For example, if we collected age groups such as "18–24", "25–38", "39–55", "56–70", and "70+", we’re now working with labels that have a built-in order where "18–24" is younger than "25–38", and so on.

Another example might be asking people whether they like ice cream, with options like "strongly agree", "agree", "neutral", "disagree", and "strongly disagree". In this case, the order of the options has meaning, even if the distance between them isn’t equal.

Take a moment to check your understanding with the exercises below.

## Exercise Checkpoint 1

```{r, echo = FALSE}
library(webexercises)
```

1.  **Describe the following variables:**

    a\. Eye colour is a `r mcq(c("discrete numerical variable", "continuous numerical variable", "ordinal categorical variable", answer = "nominal categorical variable"))`

    b\. Temperature is a `r mcq(c("discrete numerical variable", "ordinal categorical variable", "nominal categorical variable", answer = "continuous numerical variable"))`

    c\. The number of heads in a series of coin flips is a `r mcq(c( "continuous numerical variable", "ordinal categorical variable", "nominal categorical variable", answer = "discrete numerical variable"))`

    d\. Education level is a `r mcq(c("discrete numerical variable", "continuous numerical variable", "nominal categorical variable", answer = "ordinal categorical variable"))`

    e\. Postal Code is a `r mcq(c("discrete numerical variable", "continuous numerical variable", "ordinal categorical variable", answer = "nominal categorical variable"))`

    f\. Income salary is a `r mcq(c("discrete numerical variable", "ordinal categorical variable", "nominal categorical variable", answer = "continuous numerical variable"))`

    g\. Number of calls per day is a `r mcq(c("discrete numerical variable", "continuous numerical variable", "ordinal categorical variable", "nominal categorical variable", answer = "discrete numerical variable"))`

## Visually Representing Data

Now that we’ve covered the main types of variables, we can start thinking about how to visually represent data.

First, we will need to load in some data into R. We will be working with the `palmerpenguins` package. The `palmerpenguins` package is a dataset collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pallter.marine.rutgers.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/).

If you have not used this package before, you will need to install it first:

```{r, eval=FALSE}
install.packages("palmerpenguins")
```

Once installed, you will need to load the package into R.

```{r, warning = FALSE, echo=TRUE, message=FALSE}
library(palmerpenguins)
```

Now that we have the package installed, let’s take a look at the variables in the dataset using `str()`.

```{r}
str(penguins)
```

You might notice that some values are listed as `NA`. This means there are missing observations. To check for missing values, we can use:

```{r}
any(is.na(penguins))
```

Since this returns `TRUE`, we know there are missing values in the dataset. Let's find out how many:

```{r}
sum(is.na(penguins))
```

There are 19 missing observations. This can cause problems when generating numerical summaries or running statistical tests. At this stage, to handle this, we can remove the missing values. It’s always best practice to store your cleaned data in a new dataset and leave the original untouched. **Note that you should always be careful when removing observations, as even partially filled observations can still provide valuable information.**

```{r}
clean_penguins <- na.omit(penguins)
```

We can check the structure again to make sure everything looks good:

```{r}
str(clean_penguins)
```

We can see that there are eight variables. Three of these are categorical: `species`, `island`, and `sex`. These are stored as factors, where the *levels* represent the number of distinct labels or groups within each variable. For example, `sex` has two levels: `male` and `female`.

You'll notice that there are four numerical variables: `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `year`. If your instinct is that `year` doesn't quite fit as right as a numerical variable, you would be right! It is better to treat `year` as a categorical variable. To convert it, we can use:

```{r}
clean_penguins$year <- as.factor(clean_penguins$year)
```

Now if we run `str` again, we can see that `year` is now correctly stored as a categorical variable with three levels.

```{r}
str(clean_penguins)
```

If you're interested in seeing the unique groupings or labels within a categorical variable, you can use the `unique()` function.

```{r}
unique(clean_penguins$year)
```

This shows that the data was collected over the years 2007, 2008, and 2009.

You can also see from the structure output `str()` that the `penguins` dataset is stored as a 344 x 8 tibble, which is a type of data frame in R. This means that there are 344 rows (observations) and 8 columns (variables).

We can confirm this using the `dim()` function.

```{r}
dim(clean_penguins)
```

If you wish to have a closer look at inspecting the `penguins` dataset, you can use the `View()` function to open it in a new tab on RStudio.

```{r, eval=FALSE}
View(penguins)
View(clean_penguins)
```

Now that we’ve inspected the dataset, we can start thinking about how to visually represent and understand the data.

We can describe data using numerical summaries and graphs. Numerical summaries are statistical measures that capture key aspects of a dataset. The way we choose to present these summaries and the graphs that accompany them depends on the type of variable we are working with.

Let’s start with numerical summaries.

For **numerical variables**, whether continuous or discrete, we can include:

-   mean

-   standard deviation

-   median

-   minimum

-   maximum

-   Interquartile Range (IQR)

**IMPORTANT NOTE:** When reporting descriptive statistics, always provide appropriate context. If you report a mean, include the standard deviation, and vice versa. If you report a median, include the interquartile range (IQR), and vice versa. Likewise, if you report a maximum value, always include the minimum to give full perspective.

Some helpful R functions for generating summary statistics:

```{r, warning = FALSE, echo = FALSE}
library(gt)

data <- data.frame(
  "Summary Statistic" = c(
    "Mean",
    "Standard Deviation",
    "Median",
    "Minimum",
    "Maximum",
    "Interquartile Range",
    "First quartile",
    "Third quartile",
    "Five point summary"
  ),
  "Command" = c(
    "mean(...)",
    "sd(...)",
    "median(...)",
    "min(...)",
    "max(...)",
    "IQR(...)",
    "quantile(..., 0.25)",
    "quantile(..., 0.75)",
    "summary(...)"
  ),
  check.names = FALSE
)

gt(data)


```

Let's take a look at calculating descriptive statistics for `body_mass_g`, which describes the body mass of each penguin in grams.

Now we can attach the cleaned `penguins` dataset. Attaching the dataset allows us to call variables directly by name. If we don’t attach it, we can still access variables, but we need to use the full format (for example, `clean_penguins$sex`).

```{r, message=F}
# Attach cleaned dataset 
attach(clean_penguins)
```

```{r}
# Let's first calculate the mean and round it to one decimal place
round(mean(body_mass_g),1)

# Now we can calculate the standard deviation
round(sd(body_mass_g),1)
```

The average body mass of penguins is **4207.1 g**, with a standard deviation of **805.2 g**.

```{r}
# calculate the minimum
min(body_mass_g)

# calculate the maximum
max(body_mass_g)
```

The minimum observed body mass is **2700 g**, and the maximum observed body mass is **6300 g**.

```{r}
# calculate median
median(body_mass_g)

# calculate IQR
## first quantile
quantile(body_mass_g, 0.25)
## third quantile 
quantile(body_mass_g, 0.75)
```

The median body mass is **4050 g**, with an interquartile range from **3350 g to 4775 g.**

The `summary()` function lets us quickly gather all of this information at once.

```{r}
# Calculate 5-point summary
summary(body_mass_g)
```

Now that we’ve explored numerical summaries, let’s move on to graphing.

When you have one numerical variable, whether discrete or continuous, a **histogram** is a great way to visualise its distribution.

We’ll use the `ggplot2` package for this. If you haven’t used it before, you’ll need to install it first.

```{r, eval=FALSE}
# Install the package
install.packages(ggplot2)
```

Once installed, you can load the package and start plotting.

```{r, message = FALSE, warning = FALSE}
# Load the library
library(ggplot2)

# We can plot the histogram
ggplot(clean_penguins, aes(x=body_mass_g))+
  geom_histogram()
         
```

The histogram looks a bit plain without any color, and the x-axis label isn’t very informative. Let’s add some color, fix the labels, and give it a clean white background.

```{r}
# We can plot the histogram
ggplot(clean_penguins, aes(x=body_mass_g))+
  geom_histogram(fill="orangered", bins=30) +
  labs(x="Body Mass of Penguins (g)")+
  theme_bw()
```

What if we’re interested in how body mass relates to flipper length? In this case, we’re working with two continuous numerical variables, so a **scatterplot** would be a good choice.

```{r}
ggplot(clean_penguins, aes(x=body_mass_g, y=flipper_length_mm))+
  geom_point()
```

Like we did with the histogram, we can add color, change the shape of the points, and include informative labels.

```{r}
ggplot(clean_penguins, aes(x=body_mass_g, y=flipper_length_mm))+
  geom_point(fill="aquamarine", col="aquamarine4", pch=21, size=2.5)+
  labs(x="Body Mass of Penguins (g)", y="Flipper Length of Penguins (mm)")+
  theme_bw()
```

We can see that as the body mass of a penguin increases, its flipper length also tends to increase. This suggests a *strong positive linear* relationship.

Now, let’s look at a case where we have one categorical variable and one numerical variable. A **boxplot** is a great choice here. Let’s see how body mass differs between penguin species.

```{r}
ggplot(clean_penguins, aes(x=species, y=body_mass_g))+
  geom_boxplot(fill = c("coral2", "chartreuse2", "cadetblue2"),
               col = c("coral3", "chartreuse4", "cadetblue4")) +
  labs(x="Species of Penguin", y="Body Mass (g)")
```

We can see that Gentoo penguins have a noticeably higher median body mass compared to Chinstrap and Adelie penguins. The interquartile range (the box) for the Gentoo species does not overlap with the others, suggesting there may be a significant difference between them.

Since we’re now considering species, let’s also take a look at how many penguins were observed in each species. Because we’re dealing with a categorical variable, a **bar plot** is appropriate.

```{r}
ggplot(clean_penguins, aes(x = species)) +
  geom_bar(fill = c("coral2", "chartreuse2", "cadetblue2"),
               col = c("coral3", "chartreuse4", "cadetblue4")) +
  labs(x = "Penguin Species", y = "Count") +
  theme_bw()

```

We can also explore proportions to understand how each species contributes to the total sample.

```{r}
ggplot(data=clean_penguins, aes(x=species))+
  geom_bar(fill = c("coral2", "chartreuse2", "cadetblue2"),
               col = c("coral3", "chartreuse4", "cadetblue4"), 
           aes(y=after_stat(prop), group=1))+
  labs(x="Penguin Species", y="Proportion")+
  theme_bw()
```

Adelie penguins make up a much larger proportion of the sample, while Chinstrap penguins account for the smallest. It’s important to consider how such imbalances in your dataset could influence your interpretation.

We can also generate a **mosaic plot** for species. To do so, we will need to install `ggmosaic`.

```{r, eval=FALSE}
install.packages("ggmosaic")
```

```{r, message=FALSE, warning = FALSE}
library(ggmosaic)

ggplot(data=clean_penguins, aes(col=species))+
  geom_mosaic(aes(x=product(species)), fill=c("coral2", "chartreuse2", "cadetblue2"))+
  theme_bw() 
```

The wider the tile, the greater the proportion that species contributes. Adelie has the widest tile, which confirms it makes up the largest share of the sample.

What if we want to look at penguin species by the island where they were observed? We are now dealing with two categorical variables. Mosaic plots work well here too.

```{r, warning = FALSE}
ggplot(clean_penguins) +
  geom_mosaic(aes(x = product(island, species)), 
              fill = rep(c("coral2", "chartreuse2", "cadetblue2"),3)) +
  theme_bw()
```

From this plot you can see:

-   Adelie appear on Biscoe, Dream, and Torgersen in roughly similar counts.

-   Chinstrap are observed only on Dream.

-   Gentoo are observed only on Biscoe.

These patterns help explain why Adelie dominate overall, and they also show that species presence varies by island.

## Exercise Checkpoint 2 

1.  Now that you are more comfortable generating plots and summary statistics in R, it is a good time to test your skills on a new dataset, the built-in **iris** dataset. This dataset is already preloaded in RStudio, so there is no need to import anything.

    To make sure it is available, you can view it by running

    ```{r, eval=FALSE}
    View(iris)
    ```

    This will open the dataset in a new tab in RStudio so you can explore its structure and variables before starting your analysis.

    a\. How many variables are in the `iris` dataset? `r fitb(5)`

    b\. How many observations are in the `iris` dataset? `r fitb(150)`

    c\. How many categorical variables are in the `iris` dataset? `r fitb(1)`

    d\. How many numerical variables are in the `iris` dataset? `r fitb(4)`

    e\. How many species are in the `iris` dataset? `r fitb(3)`

    f\. What are the names of the three species in the `iris` dataset? Please list them in **alphabetical order**, separated by commas (for example: `x, y, z`) `r fitb("setosa, versicolor, virginica", ignore_case = TRUE, ignore_ws = TRUE, regex=TRUE)`

    g\. Is `Species` nominal or ordinal? `r fitb("nominal", ignore_case = TRUE, ignore_ws = TRUE, regex=TRUE)`

    h\. Is `Sepal.Length` continuous or discrete? `r fitb("continuous", ignore_case = TRUE, ignore_ws = TRUE, regex=TRUE)`

    i\. What is the mean sepal width of iris varieties rounded to one decimal place? `r fitb(3.1)`

    j\. What is an appropriate plot to visualise petal width against species? `r fitb("boxplot", ignore_case = TRUE, ignore_ws = TRUE, regex=TRUE)`

    k\. What is an appropriate plot to visualise petal width against sepal width? `r fitb("scatterplot", ignore_case = TRUE, ignore_ws = TRUE, regex=TRUE)`

2.  
